---
title: "Final Project Code - STAT 158"
author: Group 10 - Clara Reckhorn, Gabriella Rachal, Jocelyne Perez
format: typst
---

```{r}
library(tidyverse)
```

```{r}
###############
# Reading in and formatting data
###############

data <- read_csv("data/data.csv")

data

power_data <- data |>
  mutate(
    Y = di,
    treatment = factor(xi),
    block = factor(block)
  ) |>
  select(treatment, Y, block)

power_data
```

```{r}
treatment_map <- data.frame(
  treatment          = factor(1:8),
  source_credibility = factor(c("high","high","high","high","low","low","low","low")),
  color              = factor(c("yes","yes","no","no","yes","yes","no","no")),
  headline_framing   = factor(c("sensationalist","neutral","sensationalist","neutral",
                                "sensationalist","neutral","sensationalist","neutral"))
)

# Joining the power_data back in to match it to treatment conditions so we can do 2^3 factorial design
power_data_full <- power_data |>
  left_join(treatment_map, by = "treatment") |>
  select(source_credibility, color, headline_framing, block, Y)

power_data_full
```

### Complete Block Design Example

Step 1: Linear Model and Data Generating Process

```{r}
# political affiliation level is the block
# each participant sees exactly 1 graph (1 treatment combination)

# source cred: 0 = low, 1 = high
# color: 0 = no, 1 = yes
# headline framing: 0 = neutral, 1 = sensationalist


dry_run <- data.frame(
  unit               = 1:16,
  block              = rep(c("democrat", "republican"), each = 8),
  source_credibility = rep(c("high", "high", "high", "high",
                              "low",  "low",  "low",  "low"), 2),
  color              = rep(c("yes", "yes", "no",  "no",
                              "yes", "yes", "no",  "no"), 2),
  headline_framing   = rep(c("sensationalist", "neutral",
                              "sensationalist", "neutral",
                              "sensationalist", "neutral",
                              "sensationalist", "neutral"), 2),
  response           = c(6, 7, 5, 5, 4, 4, 3, 4,   # democrat block
                         5, 6, 4, 3, 5, 3, 2, 4)    # republican block
)

```

```{r}
# Write out the linear model as an R function

set.seed(42)

generate_Y <- function(mu, alpha, gamma, delta, beta, 
                       alpha_gamma, alpha_delta, gamma_delta, alpha_gamma_delta,
                       sigma_sq, j, k, l, i, m) {
  
  n <- length(m)
  
  epsilon <- rnorm(n, mean = 0, sd = sqrt(sigma_sq))
  
  ag  <- mapply(function(jj, kk) alpha_gamma[jj, kk], j, k)
  ad  <- mapply(function(jj, ll) alpha_delta[jj, ll], j, l)
  gd  <- mapply(function(kk, ll) gamma_delta[kk, ll], k, l)
  agd <- mapply(function(jj, kk, ll) alpha_gamma_delta[jj, kk, ll], j, k, l)
  
  Y <- mu + beta[i] + alpha[j] + gamma[k] + delta[l] + ag + ad + gd + agd + epsilon
  
  return(Y)
}
```

Step 2: Choosing params

```{r}
# Defining parameter values

# Define parameter values
mu <- 0


alpha <- c(-0.55, 0.55)      # source credibility: ~1.5 point difference
gamma <- c(-0.45, 0.45)      # color: ~0.9 point difference
delta <- c(-0.55, 0.55)      # headline framing: ~1.1 point difference
beta  <- c(-0.10, 0.10)      # block   

mu <- mean(power_data_full$Y)

sigma_sq <- var(power_data_full$Y)
n <- 16


alpha_gamma       <- matrix(c(0.15, -0.15, -0.15,  0.15), nrow = 2, ncol = 2)
alpha_delta       <- matrix(c(0.10, -0.10, -0.10,  0.10), nrow = 2, ncol = 2)
gamma_delta       <- matrix(c(0.08, -0.08, -0.08,  0.08), nrow = 2, ncol = 2)
alpha_gamma_delta <- array(c(0.05, -0.05, -0.05,  0.05,
                             -0.05,  0.05,  0.05, -0.05), dim = c(2, 2, 2))
```

## Our Estimate for Sigma Squared

```{r}
sigma_sq
```

Step 3: Simulating a dataset and Inference


```{r}
n_per_cell <- 3
i <- as.integer(power_data_full$block)
j <- as.integer(power_data_full$source_credibility)  # high=1, low=2
k <- as.integer(power_data_full$color)               # no=1, yes=2
l <- as.integer(power_data_full$headline_framing)    # neutral=1, sensationalist=2
m <- 1:nrow(power_data_full)                         # n_total = 50

n_total <- length(m)

Y <- generate_Y(mu, alpha, gamma, delta, beta,
                alpha_gamma, alpha_delta, gamma_delta, alpha_gamma_delta,
                sigma_sq, j, k, l, i, m)

dat <- data.frame(
  Y                  = Y,
  block              = factor(i),
  source_credibility = factor(j),
  color              = factor(k),
  headline_framing   = factor(l)
)

fit <- lm(Y ~ source_credibility * color * headline_framing + block, data = dat)
aov_tab <- anova(fit)

###############
# ANOVA (Table 2 & 3 in text)
###############

aov_tab
```

```{r}
# Fit linear model
fit <- lm(Y ~ source_credibility * color * headline_framing + block, data = dat)

# Extract ANOVA table
aov_tab <- anova(fit)

# Get p-value for source credibility (main effect of interest)
p_val <- aov_tab["source_credibility", "Pr(>F)"]

# Convert to rejection indicator
reject <- as.integer(p_val < 0.05)
reject
```

Step 4: Iterating to calculate Power


```{r}
one_sim <- function(mu, alpha, gamma, delta, beta,
                    alpha_gamma, alpha_delta, gamma_delta, alpha_gamma_delta,
                    sigma_sq, i, j, k, l, m) {
  
  Y <- generate_Y(mu, alpha, gamma, delta, beta,
                  alpha_gamma, alpha_delta, gamma_delta, alpha_gamma_delta,
                  sigma_sq, j, k, l, i, m)
  
  dat <- data.frame(
    Y                  = Y,
    block              = factor(i),
    source_credibility = factor(j),
    color              = factor(k),
    headline_framing   = factor(l)
  )
  
  fit <- lm(Y ~ source_credibility * color * headline_framing + block, data = dat)
  p_val <- anova(fit)["source_credibility", "Pr(>F)"]
  return(as.integer(p_val < 0.05))
}

# Run 1000 simulations
n_sim <- 1000
results <- replicate(n_sim, one_sim(mu, alpha, gamma, delta, beta,
                                    alpha_gamma, alpha_delta, gamma_delta, alpha_gamma_delta,
                                    sigma_sq, i, j, k, l, m))
power <- mean(results)
power
```

```{r}
# Iterate over Steps 1-3 multiple times: Turn the Monte Carlo crank!

# Number of simulations
iter <- 1000

# Run simulations
reject_vec <- replicate(
  iter,
  one_sim(mu, alpha, gamma, delta, beta,
          alpha_gamma, alpha_delta, gamma_delta, alpha_gamma_delta,
          sigma_sq, i, j, k, l, m)
)

# Estimated power
power_est <- mean(reject_vec)

power_est
```

```{r}
library(ggplot2)

# Running Monte Carlo estimate of power
running_power <- cumsum(reject_vec) / seq_along(reject_vec)

# Put into a data frame for ggplot
df_power <- data.frame(
  iter = seq_along(reject_vec),
  power = running_power
)

# Convergence plot
ggplot(df_power, aes(x = iter, y = power)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = mean(reject_vec), linetype = "dashed") +
  labs(
    title = "Monte Carlo Power Estimate Convergence",
    x = "Number of Simulations",
    y = "Estimated Power"
  ) +
  theme_minimal()
```



```{r}
# Power as a function of sample size

sample_sizes <- c(1, 2, 5, 10, 20, 30, 50)  # n per cell
n_sim <- 1000

power_by_n <- sapply(sample_sizes, function(n_per_cell) {
  
  # rebuild index vectors for this sample size
  n_total <- 16 * n_per_cell
  
  i <- rep(rep(1:2, each = 8), n_per_cell)
  j <- rep(rep(c(1,1,1,1,2,2,2,2), 2), n_per_cell)
  k <- rep(rep(c(1,1,2,2,1,1,2,2), 2), n_per_cell)
  l <- rep(rep(c(1,2,1,2,1,2,1,2), 2), n_per_cell)
  m <- 1:n_total
  
  results <- replicate(n_sim, one_sim(mu, alpha, gamma, delta, beta,
                                      alpha_gamma, alpha_delta, gamma_delta, alpha_gamma_delta,
                                      sigma_sq, i, j, k, l, m))
  mean(results)
})

# Plot power curve
df_n <- data.frame(n_per_cell = sample_sizes, power = power_by_n)

ggplot(df_n, aes(x = n_per_cell * 16, y = power)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "red") +
  labs(
    title = "Power vs. Sample Size",
    x = "Total N",
    y = "Estimated Power"
  ) +
  theme_minimal()

df_n$total_n <- df_n$n_per_cell * 16
df_n
```

With a sample size of 50, we have a power of 0.72.


```{r}
# swap this line for each factor/interaction:
p_val <- anova(fit)["source_credibility", "Pr(>F)"] # 0.798
p_val <- anova(fit)["color", "Pr(>F)"] # 0.429
p_val <- anova(fit)["headline_framing", "Pr(>F)"] # 0.715
p_val <- anova(fit)["source_credibility:color", "Pr(>F)"] # 0.085
p_val <- anova(fit)["source_credibility:headline_framing", "Pr(>F)"] # 0.07
p_val <- anova(fit)["color:headline_framing", "Pr(>F)"] # 0.072
p_val <- anova(fit)["source_credibility:color:headline_framing", "Pr(>F)"] # 0.053
```

## Power Curves

```{r}
one_sim <- function(mu, alpha, gamma, delta, beta,
                    alpha_gamma, alpha_delta, gamma_delta, alpha_gamma_delta,
                    sigma_sq, i, j, k, l, m) {
  
  Y <- generate_Y(mu, alpha, gamma, delta, beta,
                  alpha_gamma, alpha_delta, gamma_delta, alpha_gamma_delta,
                  sigma_sq, j, k, l, i, m)
  
  dat <- data.frame(
    Y                  = Y,
    block              = factor(i),
    source_credibility = factor(j),
    color              = factor(k),
    headline_framing   = factor(l)
  )
  
  fit <- lm(Y ~ source_credibility * color * headline_framing + block, data = dat)
  p_val <- anova(fit)["source_credibility:color:headline_framing", "Pr(>F)"]
  return(as.integer(p_val < 0.05))
}

# Run 1000 simulations
n_sim <- 1000
results <- replicate(n_sim, one_sim(mu, alpha, gamma, delta, beta,
                                    alpha_gamma, alpha_delta, gamma_delta, alpha_gamma_delta,
                                    sigma_sq, i, j, k, l, m))
power <- mean(results)
power
```

Based on our calculations, the factors that are statistically significant are (in order from most to least significant): source credibility (p = 0.0006714), headline framing (0.0014075), and color (0.0240717). The block (p = 0.82) and all two-way and three-way interactions are far from significance (all p > 0.25), which suggests that the three factors act independently with no meaningful interaction effects. 

However, for color, our power is only 0.429, which indicates that if a true effect exists, we only have about a 43 percent chance of detecting it. This result should be interpreted cautiously and it would have been mostly uninformative also if it had resulted as as non-significant.

Additionally, our two- and three-way interaction effects do not have a high enough power to be revealing. To increase power for these estimates of the effects, we would need a larger sample size.